PyTracer: Automatically Profiling Numerical Instabilities in Python

### 2022-05-16 IEEEtran/paper.tex
Fix typos

Numerical stability is a crucial requirement of reliable scientific data
processing. In unstable analyses, small numerical perturbations introduced by
data noise, software and hardware updates, or parallelization have potential to
introduce substantial deviations in final results, leading to potentially
different scientific conclusions, ultimately threatening the reliability of
computational analyses. To prevent this, numerical stability analyses need to be
conducted systematically, however, no practical tool currently exists to conduct
such analyses for Python programs.

We present PyTracer, a numerical stability profiling and visualization tool for
Python programs. \pytracer adopts a dynamic approach that evaluates numerical
stability using program execution traces, and is applicable to large programs
without manual intervention. The transparent nature of \pytracer is particularly
valuable in contrast to theoretical approaches based on condition numbers or
backward error analysis~\cite{higham2002accuracy}, which require
detailed modeling of the program and its numerical implementation. Similarly,
static code analysis techniques~\cite{goubault2006static} hardly scale
to large code bases, particularly in dynamically-typed languages such as Python.

\pytracer estimates the variability of floating-point variables by
combining program traces obtained with different numerical perturbations,
which requires (\add{a}\remove{1}) generating numerically-perturbed
executions, (\add{b}\remove{2}) tracing floating-point computations along
program executions, and (\add{c}\remove{3}) visualizing stability
evaluations. \add{PyTracer addresses (a) by relying on
Fuzzy~\cite{kiar2020comparing}, an existing collection of scientific
computing libraries instrumented with Monte-Carlo arithmetic through the
Verificarlo tool~\cite{verificarlo}, (b) by dynamically instrumenting
Python functions, modules, and classes through meta-programming, and (c) by
visualizing stability measures throughout program executions with an
interactive Plotly dashboard. \mbox{\pytracer} is available at
\mbox{\href{https://github.com/big-data-lab-team/pytracer}{github.com/big-data-lab-team/pytracer}}
under MIT License.} \remove{\mbox{\pytracer} addresses these requirements
with (1) a Fuzzy Python interpreter instrumented with Monte-Carlo
arithmetic, (2) a dynamic instrumentation of Python functions, modules, and
classes through meta-programming, (3) an interactive Plotly dashboard to
visualize stability measures throughout program executions. }

@@

\add{
    To detect numerical instabilities, \pytracer works in three main steps:
\begin{enumerate}
\item It instruments Python programs and their module dependencies
automatically, to save the input and output values of functions in a trace
file.
\item It executes the instrumented program several times using Monte-Carlo
Arithmetic to inject perturbations in function outputs.
\item It aggregates all trace files to compute statistics about the traced
values, and visualize them in a GUI.
\end{enumerate}
\pytracer adopts the workflow shown in Figure~\ref{fig:workflow}.
At step 1, \pytracer uses monkey patching to replace the functions called
by the application with a wrapper that saves their arguments and returned
values in a trace file. At step 2, it executes the application multiple
times with Monte-Carlo Arithmetic as implemented in the
Fuzzy~\cite{kiar2020comparing} environment. It then
aggregates the resulting trace files to compute summary statistics on
traced variable. At step 3, a Plotly dashboard provides interactive
visualizations highlighting unstable code sections.
}

@@

Python loads modules through an import mechanism triggered by the
\texttt{import} keyword that involves two objects: \texttt{finders} and
\texttt{loaders}. \texttt{Finders} search for packages (or
namespaces\footnote{Python distinguishes between \texttt{regular} and
    \texttt{namespace} packages. A regular package is a directory that contains a
    \_\_init\_\_.py file and potentially subdirectories (sub-packages) that contain
    themselves a \texttt{\_\_init\_\_.py} file and so on recursively. The package
    hierarchy follows those of the directory. Namespace packages introduced in
    Python 3.3 (\href{https://www.python.org/dev/peps/pep-0420/}{PEP 420}) do not
    contain an \_\_init\_\_.py file and allow for flexible directory structures.
    Hence, parts of the package can be located in zip files, on the network, or in
    separated directories. There are no functional differences between both types of
    packages.}) on the local storage --- such as standard packages installed through
the pip package manager --- or on a distant server. \texttt{Finders} do not load
modules. Instead, they return a specification (\texttt{spec} object) that
includes information on where to find and how to load modules. \texttt{Loaders}
create module objects, initialize them, and fill import-related attributes such
as \texttt{\_\_spec\_\_}. \texttt{Loaders} then execute modules and populate
their namespace. Finally, modules are bound to their import name in the
\texttt{sys.modules} dictionary.

@@

\pytracer's \texttt{loader} class first loads the original module, then copies
it as a new instance of the \texttt{ModuleType} class. It then calls the
appropriate instrumentation function for each attribute depending on its type
(function, class, or instances) and replaces the original module in the
\texttt{sys.modules} map with the instrumented version. Finally, it updates all
existing references to the original module contained in the global symbol table.
As a result, the application will transparently call the instrumented module.

@@

\remove{Listings~\mbox{\ref{fig:wrapper_creation}} and~\mbox{\ref{fig:generic_wrapper}} show
    \mbox{\pytracer}'s function instrumentation mechanism.}
For each function to instrument, \pytracer's \texttt{loader} class creates a
wrapper function, dynamically compiles it with the \texttt{compile} builtin
function, and substitutes it for the original module function.  Although simple,
this technique does not apply to callable class instances, i.e., class instances
\add{that can be called directly with the () operator, because they} are not of type
\texttt{FunctionType}. Indeed, applying the wrapper technique would 1) modify
the type of the class instance to \texttt{FunctionType}, which could cause
syntax and semantic errors, and 2) mask class attributes, leading to
\texttt{AttributeError} exceptions. \pytracer overrides the
\texttt{\_\_call\_\_} attribute with the wrapper function when possible to
overcome this issue. When the \texttt{\_\_call\_\_} attribute is not writable,
Pytracer does not instrument the class and returns a warning. 
\add{
    This limitation is expected to have a minor impact in practice since non-writable
    classes are usually called from pure Python classes that can be
    instrumented. Therefore, \pytracer reports information about the stability of
    non-writable classes through the instrumentation of the pure Python
    classes that use them.
}

@@

\add{
We do not expect this limitation to have a critical impact on the
applicability of \pytracer. 
In general, iterators involve simple computations that are unlikely to be
unstable. In particular, we encountered only a few iterators in our
experiments and none of them involved floating point values. Should
instabilities occur in iterators, \pytracer would detect them through the
inputs of the functions consuming these iterators. 
}

@@

Stochastic arithmetic leverages randomness to estimate numerical instabilities
coming from the use of floating-point representations. The main idea is to treat
round-off errors as random variables and characterize them statistically. Monte 
Carlo Arithmetic (MCA)~\cite{parker1997monte} uses this principle by introducing
two improvements: (i) a virtual precision parameter, allowing to simulate
reduced working precision, and (ii) different perturbation modes to introduce
perturbations on function inputs or output. MCA replaces each floating-point
number $x$ with a stochastic counterpart computed as:
\[
    inexact(x) =  x + \beta^{e_x - t}\xi
\]
where $\beta$ is the number base, $e_x$ is the magnitude of $x$, $t$ the virtual
precision and $\xi \in (-\frac{1}{2},\frac{1}{2})$ is a random uniform variable.
Virtual precision allows for the simulation of reduced working precision.
\add{In our experiments, we used MCA in Random Rounding (RR) mode, which
    simulates roundoff errors by applying function $inexact$ only to the
    output returned by a function. The Precision Bounding (PB) mode can
    identify other types of instabilities, in particular catastrophic
    cancellations, however, it is more difficult to apply to large code bases
    as it often leads to execution errors that require human intervention.}

Stochastic arithmetic quantifies numerical error using the number of significant
digits $s$ estimated among the sampled values. A common formula to determine
this number from MCA samples is presented in~\cite{parker1997monte}:
\begin{equation}
    s = -\log_{\beta}{ \left| \dfrac{\sigma}{\mu} \right|} \label{eq:sig-digits}
\end{equation}
where $\mu$ and $\sigma$ are the sample mean and standard deviation of a
variable sampled with MCA. \add{This can be seen as the ratio signal over noise.}
Sohier et al.~\cite{sohier2021confidence} recently provided a generalization of
this formula to include confidence intervals.

We enabled MCA in Python programs via Verificarlo~\cite{verificarlo}, a
clang-based compiler~\cite{lattner2008llvm} that replaces floating-point
operations by a generic call to a configurable floating-point model. Several
floating-point models are
available~\cite{chatelain2019automatic,chatelain2019outils}, also called
backends. \pytracer leverages Fuzzy~\cite{kiar2020comparing}, a collection of
software tools compiled with Verificarlo. In particular, Fuzzy provides
MCA-instrumented versions of the Python interpreter as well as the BLAS, LAPACK,
NumPy, SciPy, and scikit-learn libraries, which enables MCA for a wide range of
existing scientific Python programs. Fuzzy is available in Verificarlo's GitHub
organization at
\href{https://github.com/verificarlo/fuzzy}{\url{github.com/verificarlo/fuzzy}}.

@@

\pytracer stores traces in the pickle format, a binary format to serialize
Python objects. The main advantages of the pickle format compared to other
serialization approaches such as marshal or JSON are its ability to serialize
most Python objects and its native compression.  In addition, pickle writes
Python objects sequentially, which preserves the temporal ordering required by
subsequent trace analyses.

@@

Once traces are available, \pytracer sequentially parses them and computes the
mean, standard deviation, and the number of significant digits (as in
Equation~\ref{eq:sig-digits}) of all saved values. \pytracer saves these summary
statistics in an HDF5 file, a hierarchical format that facilitates visual
browsing since only required information are loaded when requested by the user.
This operation critically relies on the ordering of call input and output
objects in the traces. It assumes that the application is
single-threaded and that the control flow does not depend on a random state.
\add{Therefore, \mbox{\pytracer} cannot merge traces resulting from different
    control flow paths and will stop the parsing when encountering such a situation.}

@@

\subsection{Interactive visualization}
The third main component of \pytracer is its interactive trace visualizer built
with Plotly Dash~\cite{plotly}, a Python framework for web visualization. Plotly
Dash abstracts the low-level details of web applications in a highly
customizable development interface allowing to offload heavy computations to a
server if necessary. Figure~\ref{fig:visu-layout} presents the general layout of
the \pytracer visualizer, divided into five sub-components:
\begin{enumerate}
    \item Timeline view: This view displays the values of each input and output
          for a given function at a given invocation. By default, \pytracer
          shows the number of significant digits computed across the traces. The
          mean and standard deviation across perturbed samples are also
          available. The visualizer uses an upper triangle for inputs and a
          lower triangle for outputs. \pytracer internally represents all values
          as NumPy arrays. In the case of non-scalar values or types that cannot
          be trivially cast into NumPy arrays, \pytracer splits the value into
          several variables. Hence, \pytracer will transform a class containing
          $n$ floating-point parameters into $n$ variables. Each variable is
          prefixed with the original value and postfixed with the name of the
          parameter. Finally, non-scalar values (vectors or matrices) are
          represented by their average and can be inspected with the Zoom view
          (see below).
    \item Gantt chart view: This view displays the call graph as a Gantt chart
          representing the instrumented functions. Each function call is
          represented with a bar. The time overlaps reflect the caller/callee
          relation. For example, in the figure, \texttt{function1} calls
          \texttt{function2} that itself calls \texttt{function4} and
          \texttt{function5}. This view provides insights on the calling context
          for each function, which may influence numerical stability.
    \item  Source code view: This view displays a local snapshot of the source
          code file of the variable hovered in the timeline view.
    \item Zoom view: This view provide details about non-scalar values. For
          example, a matrix may have instabilities around its diagonal but not
          in other regions. The visualizer also provides additional metrics such
          as array shape, infinite norm, or condition number.
    \item  Functions list view: This view allows the user to select the
          functions plotted in the timeline view.
\end{enumerate}

@@

We evaluated \pytracer on two widely-used Python libraries:
SciPy~\cite{virtanen2020scipy}, the main library for scientific computing in
Python, and  scikit-learn~\cite{pedregosa2011scikit}, a reference
machine-learning framework that implements a wide range of state-of-the-art
algorithms. For each library, we executed the software examples provided in
their respective GitHub repositories. We traced each example five times, having
activated MCA through Fuzzy in the Python interpreter \add{as well as} BLAS, LAPACK, NumPy,
SciPy, and scikit-learn. \add{These libraries are written in large parts in a 
compiled language other than Python (such as C or Cython) and their execution is not 
perturbed by the sole use of Fuzzy Python. We chose to activate MCA in these libraries 
to provide an overall picture of numerical uncertainty in the tested examples. 
Depending on the use-case of interest, one might choose to use MCA only in 
a subset of these libraries or even just in the Python interpreter.} \add{We
    used MCA in Random Rounding (RR) mode, with} the default virtual
precision in Verificarlo, namely 24 bits for single-precision floating-point
values and 53 bits for double-precision values, which simulates machine error.
We used Python-3.8.5, NumPy-1.19.1, SciPy-1.5.4, and scikit-learn-0.23.2.

@@

            \multicolumn{2}{|c|}{}                                              & \penaltyRf                                                                & \valid                                                   & \valid & -                  \\

@@

\subsection{\pytracer is applicable to reference scientific computating libraries}

@@

Column \textit{sigbits} in Table~\ref{tab:pytracer_results_summary} summarizes
the \add{result} precision for the main outputs of SciPy and scikit-learn examples. We
measured the \add{result} precision in significant bits on the final primary example output
by taking the element-wise mean for non-scalar values and, if the example
produce\add{d multiple} outputs, the average value among them. \add{Significant bits were computed on the final program output 
and could therefore be computed even when parsing failed.} \add{We use the term ``result precision"
to denote the statistical variability measured in number of significant bits, in 
contrast with the other meanings of the word precision used in expressions
``virtual precision" or ``single/double precision'' in this paper.} The numerical
quality across all examples varied from 0 to 52, with an average of 44
significant bits. We note that only one example failed due to a
numerical error (\texttt{face\_recognition}
\remove{for RR and \texttt{Bayesian Ridge Regression} for full MCA}).

@@

\add{PyTracer supports native floating point formats as well as compound
    data types, such as complex floating point values, N-dimensional arrays
    and arbitrary Python classes.} Figures~\ref{fig:fft1D_inputs}
    and~\ref{fig:fft1D_outputs} \add{show the visualization of the
    \texttt{fft\_1D} input and output values. Results are complex values
    encoded in double precision. PyTracer allows visualizing the
    \add{result} precision of real (blue points) and imaginary (orange
    points) parts.} \add{The} top row in Figures~\ref{fig:fft1D_inputs}
    and~\ref{fig:fft1D_outputs} show the input mean value while the bottom
    row shows the standard deviation. \add{Even though Random Rounding
    (RR) was used, function inputs were perturbed since they result from
    multiple intermediate computations perturbed with RR.} The points with
    low magnitude in Figure~\ref{fig:fft1D_inputs} correspond to inputs
    near 0 when the input of $\sin$ or $\exp$ is close to $k\pi$, $k \in
    \mathbb{Z}$. Indeed, since MCA introduces noise, the result is not
    exactly 0. We can see in Figure~\ref{fig:fft1D_inputs} that the maximal
    noise introduced by MCA on the inputs is in the order of $10^{-14}$,
    two orders of
magnitude higher than the $ulp \simeq 10^-16$ for double precision. This slight
noise introduced can be interpreted as white noise on the input signal. We can
see on the bottom row of Figure~\ref{fig:fft1D_outputs} that the frequency noise
is of the same order of of magnitude as the input noise, which is expected. The
two peaks at x=38 and x=562 corresponds to the actual frequencies of the input
signal.

Figure~\ref{fig:spline2d_rr} \add{shows the results of the spline evaluation with
    \texttt{bisplev} from the \texttt{spline\_2D} example. It computes the 2D spline
    interpolation for function $z=(x+y)e^{-6(x^2+y^2)}$, with $(x,y) \in
        [-1,1]\times[-1,1]$. The result is a 2D-array that is natively supported by
    PyTracer. We observe a significant loss of \add{result} precision in some regions, which follows
    an interesting pattern that could be further investigated.}

@@

The second example compares a Bayesian Ridge Regression (BRR) to the Ordinary
Least Squares (OLS) estimator on a synthetic dataset and for one-dimensional
regression using polynomial feature expansion. Although the example does not
raise a runtime error, the regression coefficients computed from the fitting are
not significant. \pytracer traces reveal that the SVD solver used is once again the root
cause for this error. \add{Indeed, the BayesianRidge model first computes an SVD
    before iteratively updating the coefficients until reaching the convergence
    criterion. However, PyTracer shows an issue in the singular values computed.}
\add{We can see in Figure}~\ref{fig:BRR_svd} \add{that the last singular values is
    \texttt{NaN} due to numerical errors in the SVD, although the remaining
    singular values exhibit high \add{result} precision (between 44 and 50 bits).
    Surprinsingly, this \texttt{NaN} value does not raise an exception and is silently
    propagated. Indeed, the stopping criterion is never verified and the 
    optimization stops after reaching the maximal number of iterations.
    Figure}~\ref{fig:BRR_svd_fixed} \add{shows that uses
    the \texttt{gesvd} instead of the \texttt{gesdd} method removes the \texttt{NaN} value
    although the last singular values stills have a low \add{result} precision (below 0).
    According to Equation}~\ref{eq:sig-digits}, \add{a number of significant bits less than 0 means that the
    variability ($\sigma$) is higher than the average value ($\mu$)}.

\add{Similarly, for OLS, PyTracer traces show that the
    regression coefficients fitted are \texttt{NaN} values. By looking at the callgraph, we can
    see that the scipy calls the \texttt{lstsq} method from the
    \texttt{scipy.linalg} package which uses the \texttt{gesdd} LAPACK function.
    Figure}~\ref{fig:OLS_svd} \add{shows that one singular value is a \texttt{NaN} value,
    which corrupted the calculation of the coefficients. When using the \texttt{gesvd} method, 
    the \texttt{NaN} value disapears 
    (Fig.}~\ref{fig:OLS_svd_fixed}\add{), resulting in a \add{result} precision of
    42 bits in average in the coefficients. We can see that having
    access to all the values computed by the application with PyTracer allows
    the user to efficiently identify numerical issues.}

>>>>

Numerical stability is a crucial requirement of reliable scientific data
processing. In unstable analyses, small numerical perturbations introduced by
data noise, software and hardware updates, or parallelization have the potential to
introduce substantial deviations in final results, leading to potentially
different scientific conclusions, ultimately threatening the reliability of
computational analyses. To prevent this, numerical stability analyses need to be
conducted systematically, however, no practical tool currently exists to conduct
such analyses for Python programs.

We present PyTracer, a numerical stability profiling and visualization tool for
Python programs. \pytracer adopts a dynamic approach that evaluates numerical
stability using program execution traces and applies to large programs
without manual intervention. The transparent nature of \pytracer is particularly
valuable in contrast to theoretical approaches based on condition numbers or
backward error analysis~\cite{higham2002accuracy}, which require
detailed modeling of the program and its numerical implementation. Similarly,
static code analysis techniques~\cite{goubault2006static} hardly scale
to large codebases, particularly in dynamically-typed languages such as Python.

\pytracer estimates the variability of floating-point variables by
combining program traces obtained with different numerical perturbations,
which requires (\add{a}\remove{1}) generating numerically-perturbed
executions, (\add{b}\remove{2}) tracing floating-point computations along with
program executions, and (\add{c}\remove{3}) visualizing stability
evaluations. \add{PyTracer addresses (a) by relying on
    Fuzzy~\cite{kiar2020comparing}, an existing collection of scientific
    computing libraries instrumented with Monte-Carlo arithmetic through the
    Verificarlo tool~\cite{verificarlo}, (b) by dynamically instrumenting
    Python functions, modules, and classes through meta-programming, and (c) by
    visualizing stability measures throughout program executions with an
    interactive Plotly dashboard. \mbox{\pytracer} is available
    \mbox{\href{https://github.com/big-data-lab-team/pytracer}{github.com/big-data-lab-team/pytracer}}
    under MIT License.} \remove{\mbox{\pytracer} addresses these requirements
    with (1) a Fuzzy Python interpreter instrumented with Monte-Carlo
    arithmetic, (2) a dynamic instrumentation of Python functions, modules, and
    classes through meta-programming, (3) an interactive Plotly dashboard to
    visualize stability measures throughout program executions. }

@@

\add{
    To detect numerical instabilities, \pytracer works in three main steps:
    \begin{enumerate}
        \item It instruments Python programs and their module dependencies
              automatically, to save the input and output values of functions in a trace
              file.
        \item It executes the instrumented program several times using Monte-Carlo
              Arithmetic to inject perturbations in function outputs.
        \item It aggregates all trace files to compute statistics about the traced
              values, and visualize them in a GUI.
    \end{enumerate}
    \pytracer adopts the workflow shown in Figure~\ref{fig:workflow}.
    In step 1, \pytracer uses monkey patching to replace the functions called
    by the application with a wrapper that saves their arguments and returned
    values in a trace file. In step 2, it executes the application multiple
    times with Monte-Carlo Arithmetic as implemented in the
    Fuzzy~\cite{kiar2020comparing} environment. It then
    aggregates the resulting trace files to compute summary statistics on
    traced variables. In step 3, a Plotly dashboard provides interactive
    visualizations highlighting unstable code sections.
}

@@

Python loads modules through an import mechanism triggered by the
\texttt{import} keyword that involves two objects: \texttt{finders} and
\texttt{loaders}. \texttt{Finders} search for packages (or
namespaces\footnote{Python distinguishes between \texttt{regular} and
    \texttt{namespace} packages. A regular package is a directory that contains an
    \_\_init\_\_.py file and potentially subdirectories (sub-packages) that contain
    themselves an \texttt{\_\_init\_\_.py} file and so on recursively. The package
    hierarchy follows those of the directory. Namespace packages introduced in
    Python 3.3 (\href{https://www.python.org/dev/peps/pep-0420/}{PEP 420}) do not
    contain an \_\_init\_\_.py file and allow for flexible directory structures.
    Hence, parts of the package can be located in zip files, on the network, or in
    separated directories. There are no functional differences between both types of
    packages.}) on the local storage --- such as standard packages installed through
the pip package manager --- or on a distant server. \texttt{Finders} do not load
modules. Instead, they return a specification (\texttt{spec} object) that
includes information on where to find and how to load modules. \texttt{Loaders}
create module objects, initialize them, and fill import-related attributes such
as \texttt{\_\_spec\_\_}. \texttt{Loaders} then execute modules and populate
their namespace. Finally, modules are bound to their import name in the
\texttt{sys.modules} dictionary.

@@

\pytracer's \texttt{loader} class first loads the original module and then copies
it as a new instance of the \texttt{ModuleType} class. It then calls the
appropriate instrumentation function for each attribute depending on its type
(function, class, or instances) and replaces the original module in the
\texttt{sys.modules} map with the instrumented version. Finally, it updates all
existing references to the original module contained in the global symbol table.
As a result, the application will transparently call the instrumented module.

@@

\remove{Listings~\mbox{\ref{fig:wrapper_creation}} and~\mbox{\ref{fig:generic_wrapper}} show
    \mbox{\pytracer}'s function instrumentation mechanism.}
For each function to be instrumented, \pytracer's \texttt{loader} class creates a
wrapper function, dynamically compiles it with the \texttt{compile} builtin
function, and substitutes it for the original module function.  Although simple,
this technique does not apply to callable class instances, i.e., class instances
\add{that can be called directly with the () operator, because they} are not of type
\texttt{FunctionType}. Indeed, applying the wrapper technique would 1) modify
the type of the class instance to \texttt{FunctionType}, which could cause
syntax and semantic errors, and 2) mask class attributes, leading to
\texttt{AttributeError} exceptions. \pytracer overrides the
\texttt{\_\_call\_\_} attribute with the wrapper function when possible to
overcome this issue. When the \texttt{\_\_call\_\_} attribute is not writable,
Pytracer does not instrument the class and returns a warning. 
\add{
    This limitation is expected to have a minor impact in practice since non-writable
    classes are usually called from pure Python classes that can be
    instrumented. Therefore, \pytracer reports information about the stability of
    non-writable classes through the instrumentation of the pure Python
    classes that use them.
}

@@

\add{
    We do not expect this limitation to have a critical impact on the
    applicability of \pytracer. 
    In general, iterators involve simple computations that are unlikely to be
    unstable. In particular, we encountered only a few iterators in our
    experiments and none of them involved floating-point values. Should
    instabilities occur in iterators, \pytracer would detect them through the
    inputs of the functions consuming these iterators. 
}

@@

Stochastic arithmetic leverages randomness to estimate numerical instabilities
coming from the use of floating-point representations. The main idea is to treat
round-off errors as random variables and characterize them statistically. Monte 
Carlo Arithmetic (MCA)~\cite{parker1997monte} uses this principle by introducing
two improvements: (i) a virtual precision parameter, allowing for simulating
reduced working precision, and (ii) different perturbation modes to introduce
perturbations on function inputs or output. MCA replaces each floating-point
number $x$ with a stochastic counterpart computed as:
\[
    inexact(x) =  x + \beta^{e_x - t}\xi
\]
where $\beta$ is the number base, $e_x$ is the magnitude of $x$, $t$ the virtual
precision and $\xi \in (-\frac{1}{2},\frac{1}{2})$ is a random uniform variable.
Virtual precision allows for the simulation of reduced working precision.
\add{In our experiments, we used MCA in Random Rounding (RR) mode, which
    simulates roundoff errors by applying function $inexact$ only to the
    output returned by a function. The Precision Bounding (PB) mode can
    identify other types of instabilities, in particular catastrophic
    cancellations, however, it is more difficult to apply to large code bases
    as it often leads to execution errors that require human intervention.}

Stochastic arithmetic quantifies numerical error using the number of significant
digits $s$ estimated among the sampled values. A common formula to determine
this number from MCA samples is presented in~\cite{parker1997monte}:
\begin{equation}
    s = -\log_{\beta}{ \left| \dfrac{\sigma}{\mu} \right|} \label{eq:sig-digits}
\end{equation}
where $\mu$ and $\sigma$ are the sample mean and standard deviation of a
variable sampled with MCA. \add{This can be seen as the ratio of signal over noise.}
Sohier et al.~\cite{sohier2021confidence} recently provided a generalization of
this formula to include confidence intervals.

We enabled MCA in Python programs via Verificarlo~\cite{verificarlo} a
clang-based compiler~\cite{lattner2008llvm} that replaces floating-point
operations with a generic call to a configurable floating-point model. Several
floating-point models are
available~\cite{chatelain2019automatic,chatelain2019outils}, also called
backends. \pytracer leverages Fuzzy~\cite{kiar2020comparing}, a collection of
software tools compiled with Verificarlo. In particular, Fuzzy provides
MCA-instrumented versions of the Python interpreter as well as the BLAS, LAPACK,
NumPy, SciPy, and scikit-learn libraries, which enables MCA for a wide range of
existing scientific Python programs. Fuzzy is available in Verificarlo's GitHub
organization at
\href{https://github.com/verificarlo/fuzzy}{\url{github.com/verificarlo/fuzzy}}.

@@

\pytracer stores trace in the pickle format, a binary format to serialize
Python objects. The main advantages of the pickle format compared to other
serialization approaches such as marshal or JSON are its ability to serialize
most Python objects and its native compression.  In addition, pickle writes
Python objects sequentially, which preserves the temporal ordering required by
subsequent trace analyses.

@@

Once traces are available, \pytracer sequentially parses them and computes the
mean, standard deviation, and the number of significant digits (as in
Equation~\ref{eq:sig-digits}) of all saved values. \pytracer saves these summary
statistics in an HDF5 file, a hierarchical format that facilitates visual
browsing since only required information is loaded when requested by the user.
This operation critically relies on the ordering of call input and output
objects in the traces. It assumes that the application is
single-threaded and that the control flow does not depend on a random state.
\add{Therefore, \mbox{\pytracer} cannot merge traces resulting from different
    control flow paths and will stop the parsing when encountering such a situation.}

@@

\subsection{Interactive visualization}
The third main component of \pytracer is its interactive trace visualizer built
with Plotly Dash~\cite{plotly}, a Python framework for web visualization. Plotly
Dash abstracts the low-level details of web applications in a highly
customizable development interface allowing to offload heavy computations to a
server if necessary. Figure~\ref{fig:visu-layout} presents the general layout of
the \pytracer visualizer, divided into five sub-components:
\begin{enumerate}
    \item Timeline view: This view displays the values of each input and output
          for a given function at a given invocation. By default, \pytracer
          shows the number of significant digits computed across the traces. The
          mean and standard deviation across perturbed samples are also
          available. The visualizer uses an upper triangle for inputs and a
          lower triangle for outputs. \pytracer internally represents all values
          as NumPy arrays. In the case of non-scalar values or types that cannot
          be trivially cast into NumPy arrays, \pytracer splits the value into
          several variables. Hence, \pytracer will transform a class containing
          $n$ floating-point parameters into $n$ variables. Each variable is
          prefixed with the original value and postfixed with the name of the
          parameter. Finally, non-scalar values (vectors or matrices) are
          represented by their average and can be inspected with the Zoom view
          (see below).
    \item Gantt chart view: This view displays the call graph as a Gantt chart
          representing the instrumented functions. Each function call is
          represented with a bar. The time overlaps reflect the caller/callee
          relation. For example, in the figure, \texttt{function1} calls
          \texttt{function2} which itself calls \texttt{function4} and
          \texttt{function5}. This view provides insights into the calling context
          for each function, which may influence numerical stability.
    \item  Source code view: This view displays a local snapshot of the source
          code file of the variable hovered in the timeline view.
    \item Zoom view: This view provides details about non-scalar values. For
          example, a matrix may have instabilities around its diagonal but not
          in other regions. The visualizer also provides additional metrics such
          as array shape, infinite norm, or condition number.
    \item  Functions list view: This view allows the user to select the
          functions plotted in the timeline view.
\end{enumerate}

@@

We evaluated \pytracer on two widely-used Python libraries:
SciPy~\cite{virtanen2020scipy}, the main library for scientific computing in
Python, and  scikit-learn~\cite{pedregosa2011scikit}, a reference
machine-learning framework that implements a wide range of state-of-the-art
algorithms. For each library, we executed the software examples provided in
their respective GitHub repositories. We traced each example five times, having
activated MCA through Fuzzy in the Python interpreter \add{as well as} BLAS, LAPACK, NumPy,
SciPy, and scikit-learn. \add{These libraries are written in large parts in a
    compiled language other than Python (such as C or Cython) and their execution is not 
    perturbed by the sole use of Fuzzy Python. We chose to activate MCA in these libraries 
    to provide an overall picture of numerical uncertainty in the tested examples. 
    Depending on the use case of interest, one might choose to use MCA only in 
    a subset of these libraries or even just in the Python interpreter.} \add{We
    used MCA in Random Rounding (RR) mode, with} the default virtual
precision in Verificarlo, namely 24 bits for single-precision floating-point
values and 53 bits for double-precision values, which simulates machine error.
We used Python-3.8.5, NumPy-1.19.1, SciPy-1.5.4, and scikit-learn-0.23.2.

@@

            \multicolumn{2}{|c|}{}                                              & \penaltyRf                                                                & \valid                                                   & \valid & 52                 \\

@@

\subsection{\pytracer is applicable to reference scientific computing libraries}

@@

Column \textit{sigbits} in Table~\ref{tab:pytracer_results_summary} summarizes
the \add{result} precision for the main outputs of SciPy and scikit-learn examples. We
measured the \add{result} precision in significant bits on the final primary example output
by taking the element-wise mean for non-scalar values and, if the example
produce\add{d multiple} outputs, the average value among them. \add{Significant bits were computed on the final program output
    and could therefore be computed even when parsing failed.} \add{We use the term ``result precision"
    to denote the statistical variability measured in the number of significant bits, in 
    contrast with the other meanings of the word precision used in the expressions
    ``virtual precision" or ``single/double precision'' in this paper.} The numerical
quality across all examples varied from 0 to 52, with an average of 44
significant bits. We note that only one example failed due to a
numerical error (\texttt{face\_recognition}
\remove{for RR and \texttt{Bayesian Ridge Regression} for full MCA}).

@@

\add{PyTracer supports native floating-point formats as well as compound
    data types, such as complex floating-point values, N-dimensional arrays,
    and arbitrary Python classes.} Figures~\ref{fig:fft1D_inputs}
and~\ref{fig:fft1D_outputs} \add{show the visualization of the
    \texttt{fft\_1D} input and output values. Results are complex values
    encoded in double precision. PyTracer allows visualizing the
    \add{result} precision of real (blue points) and imaginary (orange
    points) parts.} \add{The} top row in Figures~\ref{fig:fft1D_inputs}
and~\ref{fig:fft1D_outputs} show the input mean value while the bottom
row shows the standard deviation. \add{Even though Random Rounding
    (RR) was used, function inputs were perturbed since they result from
    multiple intermediate computations perturbed with RR.} The points with
low magnitude in Figure~\ref{fig:fft1D_inputs} correspond to inputs
near 0 when the input of $\sin$ or $\exp$ is close to $k\pi$, $k \in
    \mathbb{Z}$. Indeed, since MCA introduces noise, the result is not
exactly 0. We can see in Figure~\ref{fig:fft1D_inputs} that the maximal
noise introduced by MCA on the inputs is in the order of $10^{-14}$,
two orders of magnitude higher than the $ulp \simeq 10^-16$ for double precision. This slight
noise introduced can be interpreted as white noise on the input signal. We can
see on the bottom row of Figure~\ref{fig:fft1D_outputs} that the frequency noise
is of the same order of magnitude as the input noise, which is expected. The
two peaks at x=38 and x=562 correspond to the actual frequencies of the input
signal.

Figure~\ref{fig:spline2d_rr} \add{shows the results of the spline evaluation with
    \texttt{bisplev} from the \texttt{spline\_2D} example. It computes the 2D spline
    interpolation for function $z=(x+y)e^{-6(x^2+y^2)}$, with $(x,y) \in
        [-1,1]\times[-1,1]$. The result is a 2D array that is natively supported by
    PyTracer. We observe a significant loss of \add{result} precision in some regions, which follows
    an interesting pattern that could be further investigated.}

@@

The second example compares a Bayesian Ridge Regression (BRR) to the Ordinary
Least Squares (OLS) estimator on a synthetic dataset and for one-dimensional
regression using polynomial feature expansion. Although the example does not
raise a runtime error, the regression coefficients computed from the fitting are
not significant. \pytracer traces reveal that the SVD solver used is once again the root
cause for this error. \add{Indeed, the BayesianRidge model first computes an SVD
    before iteratively updating the coefficients until reaching the convergence
    criterion. However, PyTracer shows an issue in the singular values computed.}
\add{We can see in Figure}~\ref{fig:BRR_svd} \add{that the last singular value is
    \texttt{NaN} due to numerical errors in the SVD, although the remaining
    singular values exhibit high \add{result} precision (between 44 and 50 bits).
    Surprisingly, this \texttt{NaN} value does not raise an exception and is silently
    propagated. Indeed, the stopping criterion is never verified and the 
    optimization stops after reaching the maximal number of iterations.
    Figure}~\ref{fig:BRR_svd_fixed} \add{shows that using
    the \texttt{gesvd} instead of the \texttt{gesdd} method removes the \texttt{NaN} value
    although the last singular values stills have a low \add{result} precision (below 0).
    According to Equation}~\ref{eq:sig-digits}, \add{a number of significant bits less than 0 means that the
    variability ($\sigma$) is higher than the average value ($\mu$)}.

\add{Similarly, for OLS, PyTracer traces show that the
    regression coefficients fitted are \texttt{NaN} values. By looking at the callgraph, we can
    see that the scipy calls the \texttt{lstsq} method from the
    \texttt{scipy.linalg} package which uses the \texttt{gesdd} LAPACK function.
    Figure}~\ref{fig:OLS_svd} \add{shows that one singular value is a \texttt{NaN} value,
    which corrupted the calculation of the coefficients. When using the \texttt{gesvd} method, 
    the \texttt{NaN} value disappears 
    (Fig.}~\ref{fig:OLS_svd_fixed}\add{), resulting in a \add{result} precision of
    42 bits on average in the coefficients. We can see that having
    access to all the values computed by the application with PyTracer allows
    the user to efficiently identify numerical issues.}
